A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-03-06 paper/paper1/ms.tex
Prose on 1D Parker wind models.

\subsection{Inapplicability of the Parker Wind model}
As mentioned, most of the excess we see resides outside the Roche Lobe,  extended emissions

\subsection{The timescale of gas motion}
The The density and ionization properties of this material only permit its observation during the 1.5 days immediately after its release.

\section{1D Parker wind models}
The dramatic leading tail of atmospheric escape suggests an underlying 6-dimensionality (3 space dimensions and 3 velocity dimensions), coupled with stellar wind \citep{2022ApJ...926..226M} may be needed to depict the problem faithfully.  MHD simulations are needed.  Here we explore one-dimensional (1D) models, which offer ease of interpretation and rapid calculation.  We employ the open-source \texttt{p-winds} code \citep{2022A&A...659A..62D}, which implements a transonic 1D Parker wind model with radiative transfer of the Helium 10830 triplet \citep{2018ApJ...855L..11O,2020A&A...636A..13L}.

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally we would have a measurement of the XUV spectrum of \hatp, but the limited facilities and distance of the source preclude these challenging observations.  Instead we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric specrtral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog \citep{2011A&A...532A...6S} with synthetic x-ray coronal spectra available.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}. The move to hotter F stars may be associated with rapid changes in the stellar wind and other atmospheric changes leading to the prospect of much lower XUV flux.  So it is not clear to what extent the \object[tau Boo]{$\tau$ Boo} x-ray spectrum resembles that of \hatp.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-7}, 10^{-5}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with a high and low radiation hardness.

@@

@@

\subsection{Physical mechanisms for atmospheric escape}
\object{HAT-P-67 b} exhibits both dramatic atmospheric escape and significant radius inflation.  Ohmic dissipation stands out as a physical mechanism that could be responsible for both of these attributes.  \citet{2011ApJ...738....1B} showed that ohmic heating acts to inflate hot Jupiters, with low mass hot Jupiters overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The ohmic heating scenario requires only on a modest planetary magnetic field and enough photoionization to ionize some modest fraction of neutral metals.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K, well matched to HAT-P-67 b's $\sim$1900 K.  The inflation timescale goes as (Eq. 20 in \citet{2011ApJ...738....1B}):
\begin{equation}
    \tau_{\textit{infl}}\sim \left(\frac{0.01}{\epsilon} \right) \left(\frac{M}{M_{J}}\right)^2 \left(\frac{R_{J}}{R}\right)^3 \left(\frac{1500 \textrm{K}}{T_\mathrm{eff}}\right)^4 \textrm{Gyr}.
\end{equation}

\noindent yielding an incredibly short 5 Myr timescale for HAT-P-67 b, assuming a typical $\epsilon\sim0.01$.  The order of magnitude of this inflation timescale is so fleetingly short that HAT-P-67 b must be firmly in the runaway stage of inflation, growing in surface area and proportionally fueling the ohmic dissipation mechanism.  This runaway Roche lobe overflow would explain the large leading tail as the byproduct of such profound mass loss as to create an orbit-filling torus of material.

@@

\subsection{Timescale for tidal synchronization}

@@

\software{  \texttt{pandas} \citep{mckinney10},
    \texttt{emcee} \citep{foreman13},
    \texttt{matplotlib} \citep{hunter07},
    \texttt{numpy} \citep{2020NumPy-Array},
    \texttt{scipy} \citep{2020SciPy-NMeth},
    \texttt{ipython} \citep{perez07},
    \texttt{seaborn} \citep{waskom14},
    \texttt{astropy} \citep{2022ApJ...935..167A},
    \texttt{muler} \citep{2022JOSS....7.4302G},
    \texttt{telfit} \citep{2014AJ....148...53G},
    \texttt{exoplanet} \citep{exoplanet:joss},
    \texttt{jupyter} \citep{Kluyver2016jupyter},
}

>>>>

\section{1D Parker wind models and mass loss rates}
The dramatic leading tail of atmospheric escape suggests that more complicated 3D hydrodymaic simulations coupled with stellar wind \citep{2022ApJ...926..226M} may be needed to depict the geometry faithfully.  Here we explore one-dimensional (1D) models, which offer ease of interpretation and rapid calculation.  We employ the open-source \texttt{p-winds} code \citep{2022A&A...659A..62D}, which implements a transonic 1D Parker wind model with radiative transfer of the Helium 10830 triplet \citep{2018ApJ...855L..11O,2020A&A...636A..13L}.

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally we would have a measurement of the XUV spectrum of \hatp, but the limited facilities and distance of the source preclude these challenging observations.  Instead we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with synthetic x-ray coronal spectra available \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with rapid changes in the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for \hatp.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} x-ray spectrum to \hatp.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-7}, 10^{-5}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with a high and low radiation hardness.

@@

The SED shows two key features.   First the corona model does not extend all the way to 2600 $\AA$ photons capable of ionizing out of the 2 $^3$S Helium metastable state.  This apparent deficit probably does not matter because the F spectral type obtains most of its NUV photons ($\lambda\sim2600\AA$) from the Wein side of the photospheric spectrum, so the hardness of the radiation stems mostly from the assumptions of the corona flux level.  The wavelength-dependent cross section for absorption (not shown) is largest just blueward of the ionization thresholds, so the differences in the spectrum between 504 $\AA$ and 1000 $\AA$ have relatively little impact on the overall ionization out of the metastable state.

Equipped with these three SEDs of differing radiation hardness, we explored different mass loss rates and exosphere temperatures.  Figure \ref{fig:pwinds} shows one example model spectrum with $\dot{M} = 7\times10^{13}$ g/s, and $T_0=9100$~K.  The model spectrum exhibits approximately the same line width and depth as our HPF observations, making it a plausible solution.  The model-dependent line profile (\emph{i.e. width and depth}) is degenerate with XUV flux, $\dot{M}$, and $T_0$.  We were able to replicate the line profile with mass loss rates as high as \censorbox{X}, assuming \censorbox{Y}, or as low as \censorbox{Z} assuming \censorbox{W}.

@@

\subsection{Physical mechanisms for atmospheric escape}
\object{HAT-P-67 b} exhibits both dramatic atmospheric escape and significant radius inflation.  Ohmic dissipation stands out as a physical mechanism that could be responsible for both of these attributes.  \citet{2011ApJ...738....1B} showed that ohmic heating acts to inflate hot Jupiters, with low mass hot Jupiters overflowing their Roche lobes and leading to evaporation on Gyr timescales.  The Ohmic heating scenario requires only on a modest planetary magnetic field ($\sim$1 G) and enough photoionization to ionize some modest fraction of neutral metals.  The ``sweet spot'' for this phenomenon appears to prefer equilibrium temperatures in the range of $1500<T_\mathrm{eq}<2000$ K \citep{2011ApJ...738....1B}, well matched to HAT-P-67 b's $\sim$1900 K.  A plausible scaling law relationship approximate to an order-of-magnitude predicts an inflation timescales (Eq. 20 in \citet{2011ApJ...738....1B}):
\begin{equation}
    \tau_{\textit{infl}}\sim \left(\frac{0.01}{\epsilon} \right) \left(\frac{M}{M_{J}}\right)^2 \left(\frac{R_{J}}{R}\right)^3 \left(\frac{1500 \textrm{K}}{T_\mathrm{eff}}\right)^4 \textrm{Gyr}.
\end{equation}

\noindent yielding an incredibly short 5 Myr timescale for \hatpb, assuming a typical $\epsilon\sim0.01$. The order of magnitude of this inflation timescale is so fleetingly short that \hatpb must be in the runaway stage of inflation, rapidly losing mass, and growing in surface area to fuel a positive feedback loop.

\subsection{Photoevaporation}
The prospect of Ohmic dissipation does not preclude the possibility that other mass loss drivers are at play.  Photoevaporation should simultaneously drive heating of the upper layers in the atmosphere, reinforcing the mass loss associated with Ohmic dissipation.  It is not clear which effect predominates.

@@

%\subsection{Timescale for tidal synchronization}

@@

\software{  \texttt{pandas} \citep{mckinney10},
    \texttt{emcee} \citep{foreman13},
    \texttt{matplotlib} \citep{hunter07},
    \texttt{numpy} \citep{2020NumPy-Array},
    \texttt{scipy} \citep{2020SciPy-NMeth},
    \texttt{ipython} \citep{perez07},
    \texttt{seaborn} \citep{waskom14},
    \texttt{astropy} \citep{2022ApJ...935..167A},
    \texttt{muler} \citep{2022JOSS....7.4302G},
    \texttt{telfit} \citep{2014AJ....148...53G},
    \texttt{exoplanet} \citep{exoplanet:joss},
    \texttt{jupyter} \citep{Kluyver2016jupyter},
    \texttt{p-winds}, \citep{2022A&A...659A..62D}
}
